Boosting the Accuracy of Finite Difference Schemes via Optimal Time Step Selection

### 2008-12-15 papers/OTSPDE/OTSPDE.tex
OTSPDE: - added keywords to manuscript
OTSPDE - fixed typo in OTSPDE paper
OTSPDE: - fixed references for shapes of markers in figure.

@@

To analyze the discretization error for (\ref{eq:burgers_1d_FD_scheme}), 
we combine~(\ref{eq:diffusion_eqn_1d_space_err}) from our analysis of 
the diffusion equation with the leading-order discretization error for 
the central difference approximation of $u_x$,
\beq
  \frac{\tu^{n}_{j+1} - \tu^{n}_{j-1}}{2 \dx}  =
  \frac{\partial \tu^n}{\px} 
  + \frac{\dx^2}{6} \frac{\partial^3 \tu^n}{\px^3} 
  + O(\dx^4),
  \label{eq:burgers_1d_ux_err}
\eeq
and the Burgers equation analogue of (\ref{eq:diffusion_eqn_1d_time_err}),
\bea
  \tu^{n+1}_j = \tu^{n}_j 
    &+& \dt \left( \nu \frac{\partial^2 \tu^n}{\px^2} 
                 - \tu^{n}_j \frac{\partial \tu^n}{\px} \right)
    \nonumber \\
    &+& \frac{\dt^2}{2} 
        \left( \nu^2 \frac{\partial^4 \tu^n}{\px^4} 
             - 2 \nu \tu^n_j\frac{\partial^3 \tu^n}{\px^3} 
             - 4 \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2} 
             + 2 \tu^n_j \left(\frac{\partial \tu^n}{\px} \right)^2
             + \left(\tu^n_j\right)^2 \frac{\partial^2 \tu^n}{\px^2} 
        \right) 
    \nonumber \\
    &+& O \left( \dt^3 \right).
    \label{eq:burgers_1d_time_err}
\eea
Together these imply that the leading order term in the local truncation error 
for~(\ref{eq:burgers_1d_FD_scheme}) is given by
\bea
  \tau^n_j &=&
      \left( \nu \frac{\partial^4 \tu^n}{\px^4} 
           - 2 \tu^n_j \frac{\partial^3 \tu^n}{\px^3} \right)
      \left[ \frac{\dx^2}{12} - \frac{\nu \dt}{2}  \right] \dt
  \nonumber \\
  &+& \frac{\dt^2}{2} 
      \left( 
           4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
         - 2 \tu^n_j \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
         - \left(\tu^n_j\right)^2 \frac{\partial^2 \tu^n}{\px^2}
      \right)
  \label{eq:burgers_1d_err_eqn}
\eea
Therefore, the optimal time step is $\dto = \dx^2/6D$ and the correction term 
is
\beq
  - \frac{\dt^2}{2} 
      \left( 
           4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
         - 2 \tu^n_j \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
         - \left(\tu^n_j\right)^2 \frac{\partial^2 \tu^n}{\px^2}
      \right)
  \label{eq:burgers_1d_corr_term}
\eeq 
By using the optimal time step and the correction term, the overall error
for the finite difference scheme~(\ref{eq:burgers_1d_FD_scheme}) is reduced
from $O(\dx^2)$ to $O(\dx^4)$.  As for the diffusion equation, we need
to use a second-order accurate discretization for the spatial derivatives
in the correction term.  
Figures~\ref{fig:burgers_1d_error} and \ref{fig:burgers_1d_solns} demonstrate 
the boost in accuracy obtained by using the optimal time step and adding the
correction term to the finite difference 
scheme~(\ref{eq:burgers_1d_FD_scheme}).

@@

\begin{figure}[tb]
\begin{center}
\scalebox{0.33}{
  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N}} 
\ \ 
\scalebox{0.33}{
  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N_roundoff}} 
\caption{$L^\infty$ error as a function of number of grid points for four
finite difference schemes that solve the fourth-order parabolic equation
(\ref{eq:4th_order_parabolic_eqn_1d}):
forward Euler with optimal time step and correction term (circles),
forward Euler with suboptimal time step and no correction term (squares),
Crank-Nicholson with a second-order stencil for the bilaplacian (diamonds), 
and Crank-Nicholson with a fourth-order stencil for the bilaplacian and
an $O(\dx^2)$ time step (triangles).
Both forward Euler schemes use a fourth-order, central finite difference 
approximation for the bilaplacian.  The boundary and ghost cells values
are set by taking values from the known analytical solution.
Note that the errors for the forward Euler with suboptimal time step and 
fourth-order Crank-Nicholson schemes lie almost directly on top of each 
other at the resolution of the figures.  
The figure on the left highlights the impact of round-off error associated 
with matrix inversion on the accuracy of the solution for the two 
Crank-Nicholson schemes.
}
\label{fig:4th_order_parabolic_eqn_1d_error}
\end{center}
\end{figure}

>>>>

% KEYWORDS
% optimal time step, finite difference schemes, 
% high-order accurate numerical methods, time dependent PDEs

@@

To analyze the discretization error for (\ref{eq:burgers_1d_FD_scheme}), 
we combine~(\ref{eq:diffusion_eqn_1d_space_err}) from our analysis of 
the diffusion equation with the leading-order discretization error for 
the central difference approximation of $u_x$,
\beq
  \frac{\tu^{n}_{j+1} - \tu^{n}_{j-1}}{2 \dx}  =
  \frac{\partial \tu^n}{\px} 
  + \frac{\dx^2}{6} \frac{\partial^3 \tu^n}{\px^3} 
  + O(\dx^4),
  \label{eq:burgers_1d_ux_err}
\eeq
and the Burgers equation analogue of (\ref{eq:diffusion_eqn_1d_time_err}),
\bea
  \tu^{n+1}_j = \tu^{n}_j 
    &+& \dt \left( \nu \frac{\partial^2 \tu^n}{\px^2} 
                 - \tu^{n}_j \frac{\partial \tu^n}{\px} \right)
    \nonumber \\
    &+& \frac{\dt^2}{2} 
        \left( \nu^2 \frac{\partial^4 \tu^n}{\px^4} 
             - 2 \nu \tu^n_j\frac{\partial^3 \tu^n}{\px^3} 
             - 4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2} 
             + 2 \tu^n_j \left(\frac{\partial \tu^n}{\px} \right)^2
             + \left(\tu^n_j\right)^2 \frac{\partial^2 \tu^n}{\px^2} 
        \right) 
    \nonumber \\
    &+& O \left( \dt^3 \right).
    \label{eq:burgers_1d_time_err}
\eea
Together these imply that the leading order term in the local truncation error 
for~(\ref{eq:burgers_1d_FD_scheme}) is given by
\bea
  \tau^n_j &=&
      \left( \nu \frac{\partial^4 \tu^n}{\px^4} 
           - 2 \tu^n_j \frac{\partial^3 \tu^n}{\px^3} \right)
      \left[ \frac{\dx^2}{12} - \frac{\nu \dt}{2}  \right] \dt
  \nonumber \\
  &+& \frac{\dt^2}{2} 
      \left( 
           4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
         - 2 \tu^n_j \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
         - \left(\tu^n_j\right)^2 \frac{\partial^2 \tu^n}{\px^2}
      \right)
  \label{eq:burgers_1d_err_eqn}
\eea
Therefore, the optimal time step is $\dto = \dx^2/6D$ and the correction term 
is
\beq
  - \frac{\dt^2}{2} 
      \left( 
           4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
         - 2 \tu^n_j \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
         - \left(\tu^n_j\right)^2 \frac{\partial^2 \tu^n}{\px^2}
      \right)
  \label{eq:burgers_1d_corr_term}
\eeq 
By using the optimal time step and the correction term, the overall error
for the finite difference scheme~(\ref{eq:burgers_1d_FD_scheme}) is reduced
from $O(\dx^2)$ to $O(\dx^4)$.  As for the diffusion equation, we need
to use a second-order accurate discretization for the spatial derivatives
in the correction term.  
Figures~\ref{fig:burgers_1d_error} and \ref{fig:burgers_1d_solns} demonstrate 
the boost in accuracy obtained by using the optimal time step and adding the
correction term to the finite difference 
scheme~(\ref{eq:burgers_1d_FD_scheme}).

@@

\begin{figure}[tb]
\begin{center}
\scalebox{0.33}{
  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N}} 
\ \ 
\scalebox{0.33}{
  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N_roundoff}} 
\caption{$L^\infty$ error as a function of number of grid points for four
finite difference schemes that solve the fourth-order parabolic equation
(\ref{eq:4th_order_parabolic_eqn_1d}):
forward Euler with optimal time step and correction term (circles),
forward Euler with suboptimal time step and no correction term (squares),
and Crank-Nicholson with a fourth-order stencil for the bilaplacian and
an $O(\dx^2)$ time step (diamonds), and Crank-Nicholson with a second-order 
stencil for the bilaplacian (triangles).
Both forward Euler schemes use a fourth-order, central finite difference 
approximation for the bilaplacian.  The boundary and ghost cells values
are set by taking values from the known analytical solution.
Note that the errors for the forward Euler with suboptimal time step and 
fourth-order Crank-Nicholson schemes lie almost directly on top of each 
other at the resolution of the figures.  
The figure on the left highlights the impact of round-off error associated 
with matrix inversion on the accuracy of the solution for the two 
Crank-Nicholson schemes.
}
\label{fig:4th_order_parabolic_eqn_1d_error}
\end{center}
\end{figure}
